\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2018-09-21 document/referee_response2.txt

# 
# ------------------- 
# Referee report 
# 
# This report presents a review of the article "The X-shooter GRB 
# afterglow legacy sample (XS-GRB)" by Selsing and collaborators. The 
# authors present and make available to the community a complete sample 
# of GRB afterglows observed with the X-Shooter spectrograph on the 
# VLT. This dataset is remarkable in several respect (>100 spectra 
# available taken over 8 years) and some of these have not be published 
# elsewhere. However, there are a number of important issues regarding 
# both the scientific measurements and their presentation which are 
# detailed below. These concerns need to be addressed before the paper 
# can be considered for publication. More details on these points 
# together with other comments are provided below. 
# 
# 
# 
# Major Concerns: 
# 
# - One aspect of these data that needs more emphasis is the 
# intervening absorbers. The data presented in this paper present a 
# remarkable number of metal and HI intervening absorbers as can be 
# judged from reading the appendix. These deserve much more attention, 
# including a table listing their redshifts, column density 
# measurements when available, metal lines detected, etc. A figure 
# with their redshift distribution would also be very welcome. In 
# addition, it would be interesting to compare the properties of this 
# population with the absorbers towards background quasars.

This paper is meant as a general introduction of the XSGRB legacy sample
and constitutes the first in a series of papers. While we certainly
agree that the issue of intervening absorbers is an interesting one, we
feel that it is best treated in a dedicated publication, which we are
already working on (de Ugarte Postigo et al. in prep.). This paper will
contain the complete list of intervening systems, as well as linelists,
redshifts, EWs, etc..

@@

# - The manuscript states that the sample is homogeneous which is a bit 
# misleading. The spectrograph used to make the follow-up is the same 
# for all afterglows, but the SNR (due to a combination of brightness 
# of the source and time of observations after the burst) and the 
# rest-frame wavelength coverage vary enormously. While remarkable in 
# many respects, it is not appropriate to qualify this sample as 
# homogeneous.

The referee is correct here. Our statement about homogeneity concerns
only the selection, not the outcome of our observations. We have
attempted to reword anywhere relevant and we added an explicit caveat on
the resulting S/N homogeneity in paragraph §2.1.

@@

# - There are a great many details provided about the data reduction, some of 
# which the reader fails to relate to the final scientific goals. For 
# example, a correction is applied to correct offsets in the 
# wavelength calibration, which presumably mostly impact the redshift 
# determination. How does the improvement translate into the redshift 
# determination accuracy? Similarly, for the telluric correction. What 
# is the impact of this on the final science? Conversely, a 
# description of the flux calibration error estimate would have been 
# welcome.

Concerning the wavelength calibration improvement, we have quantified
the magnitude of the improvement, and explained its relevance (§3.2). We
have also added a short paragraph to §3.5 explaining the impact of the
telluric correction on the inferred results.

Regarding the flux calibration accuracy, we have added a new section ($3.7). This includes

@@

# - The comparison of N(HI) distributions between intervening quasar 
# absorbers and intrinsic GRB absorbers is not direct. The measurement 
# of the column density of a system in the case of a quasar absorber 
# integrate across the whole system. Conversely, in the case of GRBs, 
# the burst event takes place at a random position in the host galaxy 
# and in fact, on average, the column density will measure over only 
# half of the system. This geometric effect should be taken into 
# account.

This is certainly true, when comparing QSO DLAs to GRB DLA. To accounted for the geometric effect in Fig.

#
# - Another concern is to do with using absorption features to measure 
# the redshift. Indeed, one cannot exclude that the absorption is not 
# at the position of host galaxy but lying in front of it. For this 
# reason, the redshifts provided should be seen as upper limits. 
#

# 
# Other Concerns: 
# 
# - the links to the actual data seem to be inactive. It is important to 
# be able to review the material which is made available to the 
# community. 
# 
# - caption of Fig. 3 is cut-out 
# 
# - Fig. 5: please label the y-axis 
# 
# - Fig. 10: caption is cut-out again 
# GRB150403A: could the continuum be lower? 
# GRB150915A: is this fit significant? 
# 
# - appendix A: refer to the spectral PSF in the title? 
# 
# - appendix B: science-verified, Paranalization: 
# perhaps avoid this ESO jargon? 
# 
# typos: 
# - pt. 4 p. 3: Paranl 
# - p. 3: exampl 
# - p. 3: sorrounding 
# - p. 9: curios 
# - there are numerous typos in the appendix which I will not review 
# here. Please make sure to double check spelling there. 
# 
# I would be happy to review an updated version of the manuscript where 
# the above comments are taken into account. 
# 
# -----------------------

### 2018-09-21 document/XSGRB_sample_rev1.tex
Addressing referee comments

\abstract{In this work we present spectra of all $\gamma$-ray burst (GRB)
	afterglows that have been promptly observed with the X-shooter spectrograph
	until \termdate. In total, we obtained spectroscopic observations of 103 individual GRBs
	observed within 48 hours of the GRB trigger. Redshifts have been measured for 97
	per cent of these, covering a redshift range from 0.059 to 7.84. Based on a set
	of observational selection criteria that minimize biases with regards to
	intrinsic properties of the GRBs, the follow-up effort has been focused on
	producing a homogeneous sample of 93 afterglow spectra for GRBs discovered by
	the {\em Swift} satellite. We here provide a public release of all the reduced
	spectra, including continuum estimates and telluric absorption corrections. For
	completeness, we also provide reductions for the 18 late-time observations of
	the underlying host galaxies. We provide an assessment of the degree of
	completeness with respect to the parent GRB population, in terms of the X-ray
	properties of the bursts in the sample and find that the sample presented here
	is representative of the full \textit{Swift} sample. We constrain the fraction
	of dark bursts to be < 28 per cent and we confirm previous results that higher
	optical darkness is correlated with increased X-ray absorption. For the 42
	bursts for which it is possible, we provide a measurement of the neutral
	hydrogen column density, increasing the total number of published H{\sc I}
	column density measurements by $\sim$ 33 per cent. This dataset provides a
	unique resource to study the ISM across cosmic time, from the local progenitor
	surroundings to the intervening universe.}

@@

@@

The first GRB followed up was GRB~090313 \citep{DeUgartePostigo2010}, observed
on the 15th of March, 2009, during the commissioning of X-shooter on the second
Unit Telescope (UT2) of the VLT . The bursts observed during the commissioning
or the science verification process (GRB~090313, GRB~090530, GRB~090809,
GRB~090926A) are not a part of the sample we use to investigate the statistical
properties of GRB afterglows, due to the different criteria for their selection. The
first burst observed after science verification and the mounting of X-shooter on UT2,
was GRB~091018, which is the first burst entering our statistically homogeneous
sample. For all bursts that fulfill our sample selection criteria, described in
Sect. \ref{samplecrit}, spectroscopic follow-up have been attempted with
X-shooter. Various conditions can affect our ability to follow up a given burst,
and a discussion of these conditions and their consequences for the sample is
included in Sect. \ref{badbursts}.

@@

For the majority of the bursts, we have observed with a slit width of 1\farc0,
0\farc9, and 0\farc9 for the UVB, VIS, and NIR-arm respectively. This sets a
lower limit on the delivered resolving power of the spectra based on the
tabulated values of the delivered resolutions, which is 4350, 7450, and 5300 for
the UVB, VIS and NIR-arm respectively\footnote{See
	\href{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/inst.html}{this URL} for the nominal spectral resolutions.}. %The slit width sets the width of the atmospheric sky lines and determines the amount of light lost due to the wavelength-dependent seeing PSF extending outside the coverage of the slit, where the width of sky-lines is always set by the slit width whereas both the delivered resolution and the slitloss changes for the better as the seeing PSF drops below the slit width. For atmospheric conditions delivering a seeing PSF with a FWHM of 0\farc9 observed with a 0\farc9 slit only 76.1 per cent of the light will enter the slit, meaning that for almost all observations a slitloss correction is required.
%We describe how slitlosses were corrected for in Sect. \ref{postproc}.
For accurate measurements involving line profiles, knowledge of the precise
instrumental resolution is required. The spectral resolution becomes better than
the nominal one, when the delivered seeing is smaller than the projected width
of the slit on the sky. We discuss how we determine the effective instrumental
resolution in Sect. \ref{resolution}.

@@

Since X-shooter is installed at the VLT Cassegrain focus, it is prone to
flexures. The flexures modify the projection of the slit on the detector with
respect to the one obtained in daytime calibration. This requires a modification
of the wavelength solution in order to correctly process the night-time data.
Part of this correction is performed by the pipeline using the frames taken
during the X-shooter Active Flexure Compensation procedure\footnote{X-shooter
	User Manual available at
	\href{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/doc.html}{ESO website}.}. The remaining offset is corrected by cross-correlating the observed sky spectrum with a synthetic sky spectrum \citep{Noll2012, Jones2013} after the continuum, estimated as the mode of all flux values, has been subtracted. To get the correct seeing point-spread-function (PSF) with which to convolve the synthetic sky, an initial refinement of the wavelength solution has been obtained by cross-correlating the observed sky with an unconvolved synthetic sky. This preliminary wavelength calibration is applied to the observed sky. The synthetic spectrum is then convolved with an increasing seeing PSF and the width that minimizes $\chi^2$ with the updated observed sky is chosen to be the effective sky-PSF. Using the synthetic sky with the matched resolution, a final wavelength calibration can then be calculated by cross-correlating the observed sky with the correctly broadened sky spectrum as a function of a velocity offset. Both a multiplicative and an additive offset to the wavelength calibration has been tested, but in terms of $\chi^2$, the model with only a multiplicative offset is preferred. The resulting offsets, which were smaller than 0.01 nm in the UVB and VIS data and smaller than 0.05 nm in the NIR spectra, but changing over short periods of time were applied to the corresponding spectra\footnote{The wavelength shifts have been extensively studied by ESO staff \href{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/doc/XS_wlc_shift_150615.pdf}{in this document}.}. Using the convolved synthetic sky, the pixels containing the brightest sky lines have been flagged as such in the bad pixel map.

@@

@@

@@

One of the objectives of our follow-up campaign is to measure the redshift
distribution for a well-defined, observationally unbiased and statistically useful
sample of GRBs. The imposed selection criteria (see Sect. \ref{samplecrit})
ensure that the GRBs entering our homogeneous sample, fairly represent the
underlying population. The redshift distribution of such a sample holds valuable
information about the occurrence of GRBs through cosmic time
\citep{Jakobsson2012, Perley2016a}.

>>>>

\abstract{In this work we present spectra of all $\gamma$-ray burst (GRB)
	afterglows that have been promptly observed with the X-shooter spectrograph
	until \termdate. In total, we obtained spectroscopic observations of 103 individual GRBs
	observed within 48 hours of the GRB trigger. Redshifts have been measured for 97
	per cent of these, covering a redshift range from 0.059 to 7.84. Based on a set
	of observational selection criteria that minimize biases with regards to
	intrinsic properties of the GRBs, the follow-up effort has been focused on
	producing a homogeneously selected sample of 93 afterglow spectra for GRBs discovered by
	the {\em Swift} satellite. We here provide a public release of all the reduced
	spectra, including continuum estimates and telluric absorption corrections. For
	completeness, we also provide reductions for the 18 late-time observations of
	the underlying host galaxies. We provide an assessment of the degree of
	completeness with respect to the parent GRB population, in terms of the X-ray
	properties of the bursts in the sample and find that the sample presented here
	is representative of the full \textit{Swift} sample. We constrain the fraction
	of dark bursts to be < 28 per cent and we confirm previous results that higher
	optical darkness is correlated with increased X-ray absorption. For the 42
	bursts for which it is possible, we provide a measurement of the neutral
	hydrogen column density, increasing the total number of published H{\sc I}
	column density measurements by $\sim$ 33 per cent. This dataset provides a
	unique resource to study the ISM across cosmic time, from the local progenitor
	surroundings to the intervening universe.}

@@

We remark, however, that while the selection of the afterglows has been carried
out in a homogeneous way, the resulting quality of the spectra is not uniform.
Individual bursts vary significantly in brightness, due to intrinsic
differences in luminosity, dust extinction, and time of observation. The covered
rest-frame wavelength range is also different for each case. Given that our
observations are targets-of-opportunity, this leads to a varying S/N, and in
particular the sensitivity limit to detecting absorption features varies greatly
from burst to burst and as a function of wavelength.

@@

The first GRB followed up was GRB~090313 \citep{DeUgartePostigo2010}, observed
on the 15th of March, 2009, during the commissioning of X-shooter on the second
Unit Telescope (UT2) of the VLT . The bursts observed during the commissioning
or the science verification process (GRB~090313, GRB~090530, GRB~090809,
GRB~090926A) are not a part of the sample we use to investigate the statistical
properties of GRB afterglows, due to the different criteria for their selection. The
first burst observed after science verification and the mounting of X-shooter on UT2,
was GRB~091018, which is the first burst entering our statistically homogeneously selected
sample. For all bursts that fulfill our sample selection criteria, described in
Sect. \ref{samplecrit}, spectroscopic follow-up have been attempted with
X-shooter. Various conditions can affect our ability to follow up a given burst,
and a discussion of these conditions and their consequences for the sample is
included in Sect. \ref{badbursts}.

@@

For the majority of the bursts, we have observed with a slit width of 1\farc0,
0\farc9, and 0\farc9 for the UVB, VIS, and NIR-arm respectively. This sets a
lower limit on the delivered resolving power of the spectra based on the
tabulated values of the delivered resolutions, which is 4350, 7450, and 5300 for
the UVB, VIS and NIR-arm respectively\footnote{See
	\href{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/inst.html}{this URL} for the nominal spectral resolutions.}. 
%We describe how slitlosses were corrected for in Sect. \ref{postproc}.
For accurate measurements involving line profiles, knowledge of the precise
instrumental resolution is required. The spectral resolution becomes better than
the nominal one, when the delivered seeing is smaller than the projected width
of the slit on the sky. We discuss how we determine the effective instrumental
resolution in Sect. \ref{resolution}.

@@

Since X-shooter is installed at the VLT Cassegrain focus, it is prone to
flexures. The flexures modify the projection of the slit on the detector with
respect to the one obtained in daytime calibration. This requires a modification
of the wavelength solution in order to correctly process the night-time data.
Part of this correction is performed by the pipeline using the frames taken
during the X-shooter Active Flexure Compensation procedure\footnote{X-shooter
	User Manual available at
	\href{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/doc.html}{ESO website}.}.

The remaining offset is corrected by cross-correlating the observed sky spectrum
with a synthetic sky spectrum \citep{Noll2012, Jones2013} after the continuum,
estimated as the mode of all flux values, has been subtracted. To get the
correct seeing point-spread-function (PSF) with which to convolve the synthetic
sky, an initial refinement of the wavelength solution has been obtained by
cross-correlating the observed sky with an unconvolved synthetic sky. This
preliminary wavelength calibration is applied to the observed sky. The synthetic
spectrum is then convolved with an increasing seeing PSF and the width that
minimizes $\chi^2$ with the updated observed sky is chosen to be the effective
sky-PSF. Using the synthetic sky with the matched resolution, a final wavelength
calibration can then be calculated by cross-correlating the observed sky with
the correctly broadened sky spectrum as a function of a velocity offset. Both a
multiplicative and an additive offset to the wavelength calibration has been
tested, but in terms of $\chi^2$, the model with only a multiplicative offset is
preferred. The resulting offsets, which were smaller than 0.01 nm in the UVB and
VIS data and smaller than 0.05 nm in the NIR spectra, but changing over short
periods of time were applied to the corresponding spectra\footnote{The
	wavelength shifts have been extensively studied by ESO staff
	\href{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/doc/XS_wlc_shift_150615.pdf}{in
		this document}.}. Using the convolved synthetic sky, the pixels containing the
brightest sky lines have been flagged as such in the bad pixel map.

The resulting offsets are generally small, on the order of $\sim 5$~km~s$^{-1}$,
and thus do not influence significantly the redshift measurements, whose
uncertainties are primarily driven by the S/N of the spectra, the line-shapes,
and the choice of ion species. Such a correction, however, can improve the
accuracy of line fits, especially when multiple components are tied together,
across a wide wavelength range, and allows a better comparison with
high-resolution data. The scale of the correction additionally serves to
solidify the accuracy of the wavelength solution.

@@

The effect of telluric correction varies greatly depending on the wavelength
region. In the red and ner-infrared areas, it allows to identify features in a
credible manner. which would be otherwise hidden among or blended with
atmospheric features. The telluric correction allows us to reliably recover the
continuum level even in regions of moderately low atmospheric transmission (see,
e.g., the A- and B-bands in Fig.~\ref{fig:spectrum}). We urge however to use
caution when deriving fitting lines in regions significantly affected by
telluric absorption.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Slitloss correction and flux calibration accuracy} \label{fluxes}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

For the majority of the observation presented in this sample, the slit width of
the target observations is less than the seeing PSF. This means that a fraction
of the incoming light is lost outside the slit. As an example, for atmospheric
conditions delivering a seeing PSF with a FWHM of 0\farc9 observed with a
0\farc9 slit, only 76.1 per cent of the light will enter the slit. This slitloss
is not corrected by the
flux calibration, as flux standard stars are always observed with a 5 \arcsec
slit. As a consequence, almost all spectra will, without slitloss correction,
underestimate the flux by some fraction. Additionally, slitlosses are wavelength
dependent due to the wavelength dependence of seeing \citep{Boyd1978} and as the
seeing changes across the slit in wavelength, so does the slitloss.

To correct for the slitlosses, we develop an algorithm that can, based on the
atmospheric conditions at time of observation and the wavelength dependence of
seeing, estimate the losses at each wavelength bin. We start with the DIMM
seeing \citep{Sarazin1990}, which measures the seeing FWHM at zenith. We correct
the zenith seeing by the airmass of the observational sightline, which gives us
the effective seeing, delivered outside the slit at 5000 \AA. We can then use
that the seeing $\sim \lambda^(1/5)$ to obtain a measure of the seeing across
the entire wavelength coverage of X-shooter. At each wavelength bin of the
spectra, we simulate a synthetic 2D seeing PSF based on a normalised, 2D Moffat
profile which we integrate over the used slit width. The inverse of the fraction
of the light that enters the slit, can then be used to correct for the estimated
slit losses at each wavelength bin.

To assess the accuracy of the flux-calibration, after the slit loss correction,
we compare a subset of afterglow spectra with photometry derived from the
acquisition camera. The acquisition camera takes a short image before
integration and this the delay between the image and the spectroscopic
integration is negligible. We select GRBs at DEC > 30~$^\circ$ to ensure
SDSS coverage, which allows good calibration of the photometric zero point,
acquisition camera magnitude brighter than 21.5 mag, and delay-time between GRB
trigger and spectroscopic observation < 15 hours. These cuts leaves 27
afterglow spectra covering the entire temporal span of the programme. Using this
set we can make a test of the final flux-calibration accuracy.

To compare the acquisition camera photometry with the spectra, we calculate the
synthetic magnitude of the spectrum, in the used passband, along with the
associated error. These two measures should be the same, modulo systematic
differences affecting the spectra - i.e. bag pixels/regions etc.

@@

One of the objectives of our follow-up campaign is to measure the redshift
distribution for a well-defined, observationally unbiased and statistically useful
sample of GRBs. The imposed selection criteria (see Sect. \ref{samplecrit})
ensure that the GRBs entering our homogeneously selected sample, fairly represent the
underlying population. The redshift distribution of such a sample holds valuable
information about the occurrence of GRBs through cosmic time
\citep{Jakobsson2012, Perley2016a}.
